Study of the morphofunctional signal of captivity on the microanatomy of a tarsal bone, the talus, in a wild ungulate, the wild boar (\textit{Sus scrofa})

### 2025-04-13 article.tex
mise à jour abstract
mise a jour abstract
date mesolithique calibrés

Ten specimens originate from Mesolithic levels at the Haut-des-Nachères archaeological site in Noyen-sur-Seine (Seine-et-Marne, France). This site is located in the Paris Basin and consists of fluvial peat deposits preserving archaeological layers from the Mesolithic to the Middle Neolithic (ca. 4 900–3 400 BCE; \cite{marinval-vigneArcheologiePaleoenvironnementNoyenSeine1993g,mordantNoyenSeineVingtcinqAns2013d}). Excavations revealed multiple Mesolithic occupations. The remains correspond to fluvial dynamics dated by \textsuperscript{14}C on wood (G. Delibrias/Gif, 1985–86) and span approximately one millennium: Locus 1 (NOY 83-84) = 7 400–8 000 BP ± 100; Locus 3 (NOY 85-87) = 6 200–7 000 BP ± 100. These dates were confirmed by lithic industry studies \cite{gueretLoutillagePremierMesolithique2013,
deseineNouveauRegardOccupations2019}.

The boars studied here come from Ensembles 2 and 3, attributed to the Middle Mesolithic (ca. 8 000–6 500 BCE) and Late/Final Mesolithic (ca. 6 500–5 200 BCE), respectively \cite{davidModelisationVegetationHolocene2014d}. Specimens from Ensemble 2 (K135 and E143-7) originate from Locus 1 and are included in formations dated by \textsuperscript{14}C (7 400–8 000 BP, or 7 190–5 970 BCE) to the Early Mesolithic, confirmed by lithic industry \cite{gueretLoutillagePremierMesolithique2013}. This locus is overlain by formations attributed to the Middle Neolithic II (ca. 4 000 BCE). Specimens from Ensemble 3 (Locus 3, 6 060–4 995 BCE) form a coherent group with well-clustered boar remains. There is no evidence of contamination from later deposits.

Chronologically, these populations predate the earliest Neolithic settlements (ca. 5 100–4 900 BCE in the Paris Basin), though no concrete traces exist at Noyen-sur-Seine. Furthemore, six specimens originante from the mesolithic levels from the site of l’abri des Cabônes in Ranchot (Jura, France)\cite{campyRanchotAbriCabones2019} and two from the levels of the same period of the site of l'abri d'Arconciel/ La Souche (Fribourg, Swiss) \cite{castelFauneMesolithiqueLabri2020}. Thus, this group represents pre-Neolithic economies. However, some Late Mesolithic specimens may show uncertainties regarding their wild/domestic origin, as wild specimens of domestic origin have been identified on Mesolithic sites (e.g., Ertebølle groups in Scandinavia coexisting with Neolithic Linear Pottery cultures for ~700 years; \cite{krause-kyoraUseDomesticatedPigs2013, evinExploringComplexityDomestication2014,rowley-conwyMesolithicDomesticPigs2014,rowley-conwyWildBoarDomestic2014}). Additionally, at the Grande-Rivoire shelter (Isère, France), a brown bear (\textit{Ursus arctos}) mandible with dental deformations suggests captivity \cite{chaixTamedBrownBear1997}. Pig husbandry only developed in Western Europe by the late 5th to early 4th millennium BCE \cite{arbogastPremiersElevagesNeolithiques1990,tressetSubstitutionSpeciesTechniques2007,larsonAncientDNAPig2007f}.

@@

\begin{figure}[H]
	\includegraphics[width=15cm]{correlation_matrix_pairwiseNA.jpg}
	\caption{Correlation matrix between quantitative variables of bone microanatomy and individual factors.\\
\textcolor{blue}{Blue color}: negative correlation; \textcolor{red}{red}: positive correlation. The darker the color and the larger the circle diameter, the stronger the correlation (positive or negative). Light colors correspond to values tending toward no correlation. \\
\textbf{Abbreviations:} \\
WBV: whole bone volume, C: bone compactness, \%Trab: proportion of trabecular bone, TC: trabecular compactness, RMaxT: maximum relative thickness of compact bone, RMeanT: mean relative thickness of compact bone. \\}
	\label{correlation_matrix}
\end{figure}

>>>>

Ten specimens originate from Mesolithic levels at the Haut-des-Nachères archaeological site in Noyen-sur-Seine (Seine-et-Marne, France). This site is located in the Paris Basin and consists of fluvial peat deposits preserving archaeological layers from the Mesolithic to the Middle Neolithic (ca. 4 900–3 400 BCE; \cite{marinval-vigneArcheologiePaleoenvironnementNoyenSeine1993g,mordantNoyenSeineVingtcinqAns2013d}). Excavations revealed multiple Mesolithic occupations. The remains correspond to fluvial dynamics dated by \textsuperscript{14}C on wood (G. Delibrias/Gif, 1985–86) and span approximately two millennia: Locus 1 (NOY 83-84) = (7 185 - 6 640) - (6 435 - 6 070) cal BC, $2\sigma$; Locus 3 (NOY 85-87) = (6 064 - 5 714) - (5 371 - 4 898) cal BC, $2\sigma$. These dates were confirmed by lithic industry studies \cite{gueretLoutillagePremierMesolithique2013,
deseineNouveauRegardOccupations2019}.

The boars studied here come from Ensembles 2 and 3, attributed to the Middle Mesolithic (ca. 8 000–6 500 BCE) and Late/Final Mesolithic (ca. 6 500–5 200 BCE), respectively \cite{davidModelisationVegetationHolocene2014d}. Specimens from Ensemble 2 (K135 and E143-7) originate from Locus 1 which corresponds to the Early Mesolithic. This locus is overlain by formations attributed to the Middle Neolithic II (ca. 4 000 BCE). Specimens from Ensemble 3 (Locus 3) form a coherent group with well-clustered boar remains. There is no evidence of contamination from later deposits.

Chronologically, these populations predate the earliest Neolithic settlements (ca. 5 100–4 900 BCE in the Paris Basin), though no concrete traces exist at Noyen-sur-Seine. Furthemore, six specimens originate Mesolithic assemblage from layer 3 at Les Cabônes rockshelter (Jura, France), which based on the detailed study of a large assemblage of microliths indicates that this main part of the layer belongs to the Middle Mesolithic (Sauveterrian technocomplex). The average AMS radiocarbon dates on bone provides an age between 8 200 and 7 300 cal BC \cite{campyRanchotAbriCabones2019,leduc2015wild}. In addition, two specimen come from the levels of the recent/final Mesolithic dated by radiocarbon from the second quarter of the 7th millennium to the crossroads between the 5th and 4th millennia cal BC of the site of l'abri d'Arconciel/La Souche (Fribourg, Switzerland) \cite{bassin2014fin,mauvilly2008abri,mauvilly2007vie}. Thus, this group represents pre-Neolithic economies. However, some Late Mesolithic specimens may show uncertainties regarding their wild/domestic origin, as wild specimens of domestic origin have been identified on Mesolithic sites (e.g., Ertebølle groups in Scandinavia coexisting with Neolithic Linear Pottery cultures for ~700 years; \cite{krause-kyoraUseDomesticatedPigs2013, evinExploringComplexityDomestication2014,rowley-conwyMesolithicDomesticPigs2014,rowley-conwyWildBoarDomestic2014}). Additionally, at the Grande-Rivoire shelter (Isère, France), a brown bear (\textit{Ursus arctos}) mandible with dental deformations suggests captivity \cite{chaixTamedBrownBear1997}. %Pig husbandry only developed in Western Europe by the late 5th to early 4th millennium BCE \cite{arbogastPremiersElevagesNeolithiques1990,tressetSubstitutionSpeciesTechniques2007,larsonAncientDNAPig2007f}.

@@

\begin{figure}[H]
	\includegraphics[width=15cm]{correlation_matrix_pairwiseNA.jpg}
	\caption{Correlation matrix between quantitative variables of bone microanatomy and individual factors.\\
\textcolor{blue}{Blue color}: negative correlation; \textcolor{red}{red}: positive correlation. The darker the color, the stronger the correlation (positive or negative). Light colors correspond to values tending toward no correlation. \\
\textbf{Abbreviations:} \\
WBV: whole bone volume, C: bone compactness, \%Trab: proportion of trabecular bone, TC: trabecular compactness, RMaxT: maximum relative thickness of compact bone, RMeanT: mean relative thickness of compact bone. \\}
	\label{correlation_matrix}
\end{figure}
